Tropical mathematics and the Lambda-calculus I

### 2023-07-31 CSL24/section1.tex
Commit

A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz. The fundamental motivation of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.

@@

In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (an aspect that will be developed in more detail in a future paper of this series).

@@

\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few very simple examples. This is in Sections \ref{section2} and \ref{section22}.

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ and $\mathrm{PCF}$. Notably, 
 we show that the functions interpreting higher-order programs, which correspond to a generalization of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments. This is in Sections \ref{section3} and \ref{sec:tls}.

>>>>

A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz. The fundamental intuition of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.

@@

In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (this last aspect will be developed in more detail in a future paper of this series).

@@

\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few simple examples. This is in Sections \ref{section2} and \ref{section22}.

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
 we show that the functions interpreting higher-order programs, %which correspond to a 
generalizations of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments. This is in Sections \ref{section3} and~\ref{sec:tls}.

### 2023-07-31 CSL24/section2bis.tex
Commit

\begin{definition}
A \emph{tropical power series}
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
If $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}

A tropical polynomial is always a piece-wise linear function since it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root. 
tps with more than one variable are defined as $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients. In Section \ref{section3} we also consider tps with \emph{infinitely many} variables.

@@

\begin{remark}\label{rmk:val_trop}
 There is a well-known relation between tropical polynomial/power series and usual polynomials/power series. 
Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.
%Notice that if $\mathrm{val}:Q\to ([0,+\infty],+,\cdot,0,1)$ is a valuation, then $\mathrm{logval}(x):Q\to ([-\infty,+\infty],\min,+,+\infty,0)$, $\mathrm{logval}(x):=-\log \mathrm{val}(x)$ is a valuation.

Now, for a (formal) power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.

Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$. %and the zero is $\infty$.
%Remember that a_i \neq 0 by def of formal power series
That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
\end{remark}

Observe that tropicalisation is \emph{not} in general an injective operation. 
This is because tps have a tendency to collapse, if not globally at least locally, onto tropical polynomials. For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ is  equivalent to the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.

>>>>

\begin{definition}
A \emph{tropical power series}
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}

In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.

A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

%Tropical polynomials and their roots are the main object of study of tropical geometry.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root. 
%Tps with more than one variable are defined as $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
%$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients. In Section \ref{section3} we also consider tps with \emph{infinitely many} variables.

@@

\begin{remark}\label{rmk:val_trop}
 There is a well-known relation between tropical polynomial/power series and usual polynomials/power series. 
Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
For example, the valuation which gives $1\in L$ on all the invertibles of $Q$ and $0$ otherwise, is called the \emph{trivial valuation} $\mathrm{val}_1$.
%Notice that if $\mathrm{val}:Q\to ([0,+\infty],+,\cdot,0,1)$ is a valuation, then $\mathrm{logval}(x):Q\to ([-\infty,+\infty],\min,+,+\infty,0)$, $\mathrm{logval}(x):=-\log \mathrm{val}(x)$ is a valuation.

Now, for a power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
%Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$. %and the zero is $\infty$.
%Remember that a_i \neq 0 by def of formal power series
%That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
\end{remark}

%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon}, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

### 2023-07-31 CSL24/section5bis.tex
Commit

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential constructor} $\Der[\_,\_]$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$). 
%Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
%
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
%The models of $\STDLC$ are the 
%cartesian \emph{closed} differential categories ($CC\partial C$), 
%which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}. 
The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
the idea is that, in $\STDLC$, one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{k}[M,N^k]$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$, and doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion is just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

@@

At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

>>>>

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential constructor} $\Der[\_,\_]$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once.
This is also why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$. 
In particular, the application of $\Der[M,N]$ to an ``error term'' $0$ ensures that $M$ will use $N$ exactly once (we say \emph{linearly}).
%Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
%
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
%The models of $\STDLC$ are the 
%cartesian \emph{closed} differential categories ($CC\partial C$), 
%which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.

The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
the idea is that one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{k}[M,N^k]0$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$; 
doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion becomes just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

@@

At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, %i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
due to its strict relations with the Taylor expansion of programs.
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.
